---
title: "`FeatSeekR` user guide"
author: 
- name: Tuemay Capraz
  affiliation: European Molecular Biology Laboratory, Heidelberg
  email: tuemay.capraz@embl.de
package: FeatSeekR
date: "`r Sys.Date()`"
output:  
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{`FeatSeekR` user guide}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignettePackage{FeatSeekR-vignette}
    %\VignetteEncoding{UTF-8}
---



```{r setup, message=FALSE}
library(FeatSeekR)
library(DmelSGI)
library(pheatmap)
```


# Introduction

A fundamental step in many analyses of high-dimensional data is dimension 
reduction. Feature selection is one approach to dimension reduction whose 
strengths include interpretability, conceptual simplicity, transferability 
and modularity.
Here, we introduce the `FeatSeekR` algorithm, which selects features based on 
the consistency of their signal across replicates and their non-redundancy.
It takes a 3 dimensional array (samples x features x replicates) 
or a list of replicate matrices/dataframes (samples x features)  as input 
and returns a dataframe storing the selected and ranked features and 
their replicate correlations.

# Installation

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("FeatSeekR")
```

# Feature selection on simulated data

Here we simulate a data set with features generated by orthogonal latent factors.
Features derived from the same latent factor are highly redundant and form
distinct clusters.

```{r simulate data}
set.seed(42)
# simulate data with 500 samples, 3 replicates and 5 latent factors 
# generating 50 features
samples <- 500
latent_factors <- 5
replicates <- 3

data <- FeatSeekR::simData(samples=samples,latent_factors =latent_factors,
                replicates = replicates)
data.mean = apply(data, c(1,2), mean)
# calculate the feature correlation 
cor <- stats::cor(data.mean, use = "pairwise.complete.obs")

range <- max(abs(cor))
# plot a heatmap of the features and color features according to their 
# generating latent factors
anno <- data.frame(Latent_factor = as.factor(rep(1:5, each=10)))
rownames(anno) <- dimnames(data)[[2]]
colors        <- c("red", "blue", "darkorange", "darkgreen", "black")
names(colors) <- c("1", "2", "3", "4", "5")
anno_colors <- list(Latent_factor = colors)
pheatmap::pheatmap(cor, treeheight_row = 0 , treeheight_col = 0, 
         show_rownames = FALSE, show_colnames = FALSE,
         breaks = seq(-range, range, length.out = 100), cellwidth = 6, 
         cellheight = 6, annotation_col = anno, annotation_colors = anno_colors, 
         fontsize = 8)
```
As expected, the features cluster together according to their underlying latent
factor.
We now run `FeatSeekR` to rank the features based on their uniqueness and
reproducibility.

```{r plot top 5}
# select the top 5 features
res <- FeatSeek(data, max_features=5)

# plot a heatmap of the top 5 selected features 
FeatSeekR::plotSelectedFeatures(data, res)
```

The top 5 selected features are each from different latent factor and low
correlated.

# Selecting image features from the `DmelSGI` package

Here we use `FeatSeekR` to rapidly identify unique features with reproducible 
signal between measurements in an image dataset from the `DmelSGI` package. 
The authors of `DmelSGI` performed combinatorial gene knock-outs using siRNA, 
followed by imaging of the cells. The resulting images were segmented and 
features were extracted using the `EBImage` package. 

```{r, load_data}
# load data
data("subSampleForStabilitySelection", package="DmelSGI")
data <- subSampleForStabilitySelection$D
# make sure dimensions are samples x features x replicates
data <- aperm(data, c(1,3,2))
# set feature names
dimnames(data)[[2]] <- subSampleForStabilitySelection$phenotype
dim(data)
```

The input data is a 3 dimensional array with 3000 samples, 162 features and 2
replicates.

```{r plot data}
# plot correlation matrix of the first 100 features of one of the replicates
cor_mat <- cor(data[,1:100,1])
pheatmap::pheatmap(cor_mat, show_rownames = FALSE, show_colnames = FALSE,
    treeheight_row = 0, treeheight_col = 0)
```

When plotting the correlation matrix of the features, we can see that many 
features are highly correlated and form redundant feature clusters. We apply 
FeatSeek to identify unique features with high replicate consistency.

```{r select_features}
# select up to 50 features
max_features <- 50
res <- FeatSeekR::FeatSeek(data, max_features = max_features)
```

We can further inspect the selected features by plotting the fraction of explained variance per selected feature.

```{r inspect_selection}
FeatSeekR::plotVarianceExplained(res)
```

We select the number of features that explain at least 70% of the total variance
and plot their correlation matrix.

```{r plot selection}
# get number of features which explain at least 70% of the total variance
n_feat <- min(which(res$explained_variance > 0.7))

FeatSeekR::plotSelectedFeatures(data,res, n_features = n_feat)
```


# Session Info

```{r}
sessionInfo()
```

